---
title: "Tutorial for Mixture Models"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

tutorial_options(exercise.timelimit = 60, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(beetroot)
```

## A brief introduction to mixture models

### Mixture densities

Mixture models are a type of model designed for situation where the data stem from different, unknown groups. A classic example are the eruptions times of the Old Faithful geyser in Yellowstone National Park, USA. 

```{r}
faithful %>% 
  ggplot(aes(x = eruptions)) + geom_histogram()
```

The erution times are either around 2 minutes or around 4.5 minutes. The assumption we are going to make here is that there are two different processes at work, resulting in either a shorter or longer eruption. While longer eruptions seem more common than shorter ones, both types fairly normally distributed. So we could describe the shorter eruptions through a normal density with mean 2 and the longer ones through a normal distribution with mean 4.5. 

We can describe the distribution of _all_ of the Old Faithful eruptions through a mixture of two normal densities, with the weights for the two densities representing how common short and longer eruptions are.

<!-- FIXME: add plot with two estimated densities and their mixture -->
```{r}
library(flexmix)
set.seed(3)
mm_of <- flexmix(eruptions ~ 1, k = 2, data = faithful)
#parameters(mm_of)
#prior(mm_of)
density_1 <- function(x){ 
  prior(mm_of)[1] * dnorm(x, mean = parameters(mm_of, component = 1)[1],
                          sd = parameters(mm_of, component = 1)[2])
}
density_2 <- function(x){ 
  prior(mm_of)[2] * dnorm(x, mean = parameters(mm_of, component = 2)[1],
                          sd = parameters(mm_of, component = 2)[2])
}
density_mix <- function(x){
  density_1(x) + density_2(x)
}

of <- tibble(eruptions = seq(from = 1.6, to = 5.1, by = 0.01)) %>% 
  mutate(d_1 = density_1(eruptions),
         d_2 = density_2(eruptions),
         d_mix = density_mix(eruptions))

faithful %>% 
  ggplot(aes(x = eruptions)) +
  geom_histogram(aes(y = ..density..)) +
  geom_line(aes(x = eruptions, y = d_1, col = I("red")), data = of) +
  geom_line(aes(x = eruptions, y = d_2, col = I("blue")), data = of) 
#+ geom_line(aes(x = eruptions, y = d_mix, col = I("red")), data = of)
```



### Mixture models

We know a little more about Old Faithful: we also know the waiting time between eruptions.

```{r}
faithful %>% 
  ggplot(aes(x = waiting, y = eruptions)) +
  geom_point() 
```

If we want to model eruption time based on the waiting time, we could use a simple linear model.

```{r}
faithful %>% 
  ggplot(aes(x = waiting, y = eruptions)) +  
  geom_point() + 
  geom_smooth(method = "lm")
```

However, we are assuming that the two clusters stem from two different processes. Why would we assume that the relationship between eruption and waiting time is the same for both?

If we split the data at an eruption time of 3 minutes and estimate a separate linear model for both groups, we see that the slope is steeper for the long eruptions.

```{r}
faithful %>% 
  mutate(group = factor(eruptions < 3, levels = c(TRUE, FALSE), labels = c("short", "long"))) %>% 
  ggplot(aes(x = waiting, y = eruptions)) +
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ group)
```

A mixture model extends the idea of mixture densities to models. The mixture model is a weighted sum of components, just like the mixture density. However, the components are now models instead of densities. 


### Quiz


```{r quiz_prep}
dgp <- function(reg = TRUE){
  w <- c(0.4, 0.6)
  intercept <- c(2,4)
  slope <- if (reg) c(-0.05, 0.05) else c(0, 0)
  n <- 1000
  
  df1 <- tibble(waiting = rnorm(n*w[1], 55, 10),
                eruptions = intercept[1] + slope[1] * waiting + rnorm(n*w[1]),
                cl = 1)
  df2 <- tibble(waiting = rnorm(n*w[2], 110, 10),
                eruptions = intercept[2] + slope[2] * waiting + rnorm(n*w[2]),
                cl = 2)
  bind_rows(df1, df2)
}

my_reg <- sample(x = c(0,1), size = 2, replace = FALSE) %>% as.logical()
df <- map_dfr(my_reg, ~ dgp(reg = .)) 

df %>% 
  mutate(scenario = c("A", "B")[rep(1:2, each = 1000)]) %>% 
  ggplot(aes(waiting, eruptions)) + 
  geom_point() + facet_wrap(~ scenario)

```

```{r quiz_reg}
question("In which scenario would you expect the relationship between the waiting time and the duration of the eruption to vary between clusters?",
         answer("Scenario A", correct = my_reg[1]),
         answer("Scenario B", correct = my_reg[2]),
         allow_retry = TRUE
)

```

## Fitting a mixture model with {beetroot}

The package **beetroot** implements **parsnip** bindings for mixture models so that you can use your usual tidymodels syntax. 

```{r echo=TRUE}
# remotes::install_github("hfrick/beetroot")
library(beetroot)
```
The function to specify your model is `mixture_model()` and takes the number of `components` as an argument. The beetroot package currently uses the **flexmix** package for the estimation so you need to set your engine to `"flexmix"` with the `set_engine()` function.

```{r echo=TRUE}
mm_spec <- mixture_model(components = 2) %>% 
  set_engine("flexmix")
```

You can now fit your mixture model with the `fit()` function. You can specify the relationship captured by the components via the usual formula interface, here `eruptions ~ waiting`. The dataset is specified via the `data` argument.

```{r echo=TRUE}
mm_fit <- mm_spec %>%
  fit(eruptions ~ waiting, data = faithful)
```

_Tutorial extension: add test/train split to workflow and predict on the test set._


### Exercise

```{r}
df <- df %>% 
  mutate(scenario = c("A", "B")[rep(1:2, each = 1000)]) %>% 
  filter(scenario == c("A", "B")[my_reg])
```

Fit a mixture model with 2 components for `r c("Scenario A", "Scenario B")[my_reg]` on the data in data frame `df`.

```{r fit_beetroot, exercise=TRUE, exercise.lines = 5}

```

```{r fit_beetroot-hint}
?beetroot::mixture_model
```

```{r fit_beetroot-solution}
mixture_model(components = 2) %>% 
  set_engine("flexmix") %>% 
  fit(eruptions ~ waiting, data = df)
```

```{r fit_beetroot-check}
grade_code("Well done!")
```



